---
title: "SOC/CSSS/CSDE 533 Lab 2"
author: "Sarah Kilpatrick"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
  pdf_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, echo= FALSE}
knitr::opts_chunk$set(echo = TRUE)
# knitr::opts_knit$set(root.dir = paste0("~/Dropbox/Training/CSDE533/Wi25/",
#                                        "Assignments/R/Lab 2"))
```

# Problem 1: Data and Scatterplots - 50 points

*Install and load the package `wpp2024` into your `R` environment.*

```{r install_wpp2024, eval = TRUE, echo= FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com"))
library(devtools)
options(timeout = 600)
# install_github("PPgp/wpp2024")
library(wpp2024)
library(tidyr)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(patchwork)

```

*Load the datasets `e0F1` and `e0M1`. Note:* $e_0$ *is demographic
notation for life expectancy at birth, or how many years, on average, a
person from a specific population can expect to live. These estimates
are typically made separately for two sex categories: male and female,
as there is a known gap between male and female life expectancy.*

```{r p1 setup, eval=TRUE, echo=TRUE}
data(`e0F1`)
data(`e0M1`)
```

***(a)5 points** Use several functions we've discussed to 'get to know
your data'. HINT: don't forget to use the help documentation for the
datasets/package.*

1.  *Who published these life expectancy estimates? When were they
    published?*

Reference: World Population Prospects: The 2024 Revision.
<http://population.un.org/wpp>

Details: These data come from the United Nations' annual and 5-year time
series of the life expectancy at birth (e0) for all countries of the
world as available in 2024.

"These datasets are based on estimates and projections of United
Nations, Department of Economic and Social Affairs, Population Division
(2024). The pre-1950 datasets and the 5-year datasets were compiled by
Patrick Gerland."

```{r problem 1 part a, eval=TRUE, echo=FALSE}
# summary(e0F1)
# summary(e0M1)
# head(e0F1)
# head(e0M1)
#colnames(e0F1)
?e0F1
```

2.  *What class of object is the second column of `e0F1`?*

The second column of `e0F1` is a character column called 'name'.

3.  *What is the range of female and male life expectancy for the 2023
    time period? Present your results in a table.*

```{r problem 1 part a question 3, eval=TRUE, echo=TRUE}
output = data.frame(
  Gender = c("Female", "Male"),
  Min_Life_Expectancy = c(min(e0F1$`2023`, na.rm = TRUE), min(e0M1$`2023`, na.rm = TRUE)),
  Max_Life_Expectancy = c(max(e0F1$`2023`, na.rm = TRUE), max(e0M1$`2023`, na.rm = TRUE))
)

kable(output, 
      col.names = c("Gender", "Minimum Life Expectancy in Years", "Maximum Life Expectancy in Years"),
      digits = 2,
      caption = "Range of Female/Male Life Expectancy in 2023") %>%
  kable_styling(
    font_size = 15
  )
```

***(c)10 points** Pick a country of your choice and plot both male and
female life expectancy against time.*

Pivot the data:

```{r problem 1 part c 1, eval=TRUE, echo=FALSE}
# paraguay is country code = 600
paraguay_rawfdata = e0F1[e0F1$country_code == 600, ]
paraguay_rawmdata = e0M1[e0M1$country_code == 600, ]

paraguay_long_female <- paraguay_rawfdata %>%
  pivot_longer(cols = `1950`:`2023`,
               names_to = "Year", #new column
               #where the old columns' values are now going 
               values_to = "e_0") %>% 
  select(-name, -country_code)

paraguay_long_male <- paraguay_rawmdata %>%
  pivot_longer(cols = `1950`:`2023`,
               names_to = "Year", 
               values_to = "e_0") %>% 
  select(-name, -country_code)

# paraguay_long_female
# paraguay_long_male

```

Join the data for plotting and later use:

```{r problem 1 part c 2, eval=TRUE, echo=TRUE}

paraguay_long_female = paraguay_long_female %>%
  mutate(Gender = "Female")
paraguay_long_male = paraguay_long_male %>%
  mutate(Gender = "Male")

joined_paraguay_data = bind_rows(paraguay_long_female, paraguay_long_male)

joined_paraguay_data$Year = as.numeric(joined_paraguay_data$Year)

```

Plot the data:

```{r problem 1 part c 3, eval=TRUE, echo=TRUE}
ggplot(joined_paraguay_data, aes(x = Year, y = e_0, color = Gender)) +
  geom_line(linewidth = 1) +
  ggtitle("Female and Male Life Expectancy\nin Paraguay (1950-2023)") +
  xlab("Year") +
  ylab("Life Expectancy") +
  scale_color_manual(values = c("goldenrod", "seagreen")) +
  scale_x_continuous(breaks = seq(min(joined_paraguay_data$Year), max(joined_paraguay_data$Year), by = 5)) +
geom_vline(xintercept = 2021, linetype = "dashed", color = "red", linewidth = 0.5) +
theme_minimal() +
theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 10),
    legend.title = element_blank()
  )
```

There's a sharp drop in 2021 marked with a red dashed line. This was
likely due to the COVID-19 pandemic. Paraguay's response to the pandemic
in 2020 and 2021 was to focus on sanitation, exposure tracking, and a
multi-phase return to school/work plan beginning in October 2020.

Even with the onset of vaccinations in early 2021, that year experienced
the maximum amount of deaths and infections between 2020 and when they
stopped tracking this information in March of 2022. This information was
found through the [Paraguayan Government's Department of Public Health
and Wellness](https://www.mspbs.gov.py/index.php).

1.  *Use a combination of `plot` and `points` to make a scatterplot of*
    $e_0$ *versus time. Use either distinct symbols OR distinct colors
    for each sex, and include a legend. Use the axis labels and title to
    correctly annotate your plot.*

```{r paraguay plot and points, eval=TRUE, echo=TRUE}

plot(joined_paraguay_data$Year, joined_paraguay_data$e_0, 
     type = "n",  
     xlab = "Year", ylab = "Life Expectancy", 
     main = "Female and Male Life Expectancy in Paraguay (1950-2023)",
     xaxt = "n", 
     xlim = c(min(joined_paraguay_data$Year), max(joined_paraguay_data$Year)),
     ylim = c(min(joined_paraguay_data$e_0), max(joined_paraguay_data$e_0)))

points(joined_paraguay_data$Year[joined_paraguay_data$Gender == "Female"], 
       joined_paraguay_data$e_0[joined_paraguay_data$Gender == "Female"], 
       col = "goldenrod", pch = 19)  

points(joined_paraguay_data$Year[joined_paraguay_data$Gender == "Male"], 
       joined_paraguay_data$e_0[joined_paraguay_data$Gender == "Male"], 
       col = "seagreen", pch = 19)  

# Dashed red line at x = 2021
abline(v = 2021, col = "red", lty = 2, lwd = 1) 

axis(1, at = seq(min(joined_paraguay_data$Year), max(joined_paraguay_data$Year), by = 5))

# Legend
legend("topleft", legend = c("Female", "Male"), 
       col = c("goldenrod", "seagreen"), bty = "n")


```

2.  *Use `?lines` or `help("lines")` to look up the documentation for
    the function `lines`. Use a combination of `plot` and `lines` to
    plot the male and female life expectancies from your country of
    choice. (Your plot should not contain individual points, but two
    distinct lines.) Use either color or line type to distinguish
    between the series for each sex. Include a legend and appropriate
    axis labels and a main title. Save this plot as a .pdf.*

```{r lines, eval=FALSE, echo=TRUE}
?lines
```

Save the PDF:

```{r paraguay female e0 base R to PDF, eval=TRUE, echo=TRUE}
# # Save the plot as a PDF
# pdf("paraguay_e0_lines.pdf", 
#     width = 6, height = 6)

plot(paraguay_long_female$Year, paraguay_long_female$e_0, 
     type = "l", 
     col = "goldenrod", 
     lwd = 2, 
     xlab = "Year", 
     ylab = "Life Expectancy (Years)", 
     main = "Life Expectancy at Birth in Paraguay (1950-2023)",
     ylim = range(min(paraguay_long_male$e_0), max(paraguay_long_female$e_0)))

#grid lines
grid(nx = NULL, ny = NULL, col = "lightgray")

lines(paraguay_long_male$Year, paraguay_long_male$e_0,
      col = "seagreen", 
      lwd = 2)

legend("bottomright", 
       legend = c("Female", "Male"), 
       col = c("goldenrod", "seagreen"), 
       lwd = 2)

dev.off() #turn off graphical device function off
```

Plot the lines

```{r paraguay female e0 base R, eval=TRUE, echo=TRUE}
plot(paraguay_long_female$Year, paraguay_long_female$e_0, 
     type = "l", 
     col = "goldenrod", 
     lwd = 2, 
     xlab = "Year", 
     ylab = "Life Expectancy (Years)", 
     main = "Life Expectancy at Birth in Paraguay (1950-2023)",
     ylim = range(min(paraguay_long_male$e_0), max(paraguay_long_female$e_0)))

#grid lines
grid(nx = NULL, ny = NULL, col = "lightgray")

lines(paraguay_long_male$Year, paraguay_long_male$e_0,
      col = "seagreen", 
      lwd = 2)

legend("bottomright", 
       legend = c("Female", "Male"), 
       col = c("goldenrod", "seagreen"), 
       lwd = 2)
```

***(d)20 points** Create a new data frame called `plot_data` by
subsetting `e0F1` and selecting 4 countries and the row whose
`country_code == 900` and `name = "World"`.*

I am choosing the 4 countries that had the highest frequency of tropical
cyclones in 2024. These rankings vary by source but [this
article](https://www.newsweek.com/map-shows-which-countries-get-most-hurricanes-1965603)
was reported by Newsweek, provided by the Weather Channel.

1.  China
2.  Philippines
3.  Japan
4.  Mexico

```{r filter cyclone countries, eval=TRUE, echo=FALSE}
plot_data = e0F1 %>%
  filter(country_code %in% c(156, 608, 392, 484, 900))

kable(plot_data, digits = 2) %>%
  kable_styling(
    font_size = 15
  ) %>%
  scroll_box(width = "900px", height = "500px")
```

1.  *Convert your new data.frame from **wide** format (one row per
    country) to **long** format (a row for every country/year period).
    In `tidyr` this is done using `pivot_longer`. In base R this is done
    using `reshape`. Use both `pivot_longer` and `reshape`.*

```{r pivot cyclone data long, eval=TRUE, echo=TRUE}

cyclone_data_long <- plot_data %>%
  pivot_longer(
    # Specify the columns from years 1951 to 2023
    cols = `1950`:`2023`, 
    # name of new SINGLE column for year from MANY columns
    names_to = "Year", 
    # name of new column for life expectancy values
    values_to = "e_0"      
  )

kable(cyclone_data_long, 
      col.names = c("Country Code", "Country Name", "Year", "Female Life Expectancy at Birth"),
      digits = 2,
      caption = "Long Format of plot_data: Top 4 Cyclone-Prone Nations (2024) vs. World") %>%
  kable_styling(
    font_size = 15
  ) %>%
  scroll_box(width = "900px", height = "500px")

```

2.  *How do the two functions differ in their output?*

The tidyverse function pivot_longer() outputs a tibble, while the base R
reshape() function outputs a basic R dataframe. Tibbles tend to be more
human-readable with cleaner presentation, but are not very different
from R dataframes just from appearance alone. A bigger difference is the
way the two functions differ in treating variables: the tidyverse's
output won't make the new column values into factors like the base R's
function will.

3.  *Create table showing the min, mean, and max life expectancies for
    each population in your dataset.*

```{r summarize cyclone data, eval=TRUE, echo=TRUE}
# Subsetting data for each country and population group
china_long <- subset(cyclone_data_long, name == "China")
philippines_long <- subset(cyclone_data_long, name == "Philippines")
japan_long <- subset(cyclone_data_long, name == "Japan")
mexico_long <- subset(cyclone_data_long, name == "Mexico")
world_long <- subset(cyclone_data_long, name == "World")

pop_list <- list(china_long, philippines_long, japan_long, mexico_long, world_long)

# place to store the new data
summary_e0_table <- data.frame(
  name = character(0), 
  `minimum e_0` = numeric(0), 
  `maximum e_0` = numeric(0), 
  `average e_0` = numeric(0)
)

for (populationn in pop_list) {
  pop_name <- unique(populationn$name)
  min_e0 <- min(populationn$e_0, na.rm = TRUE)
  max_e0 <- max(populationn$e_0, na.rm = TRUE)
  mean_e0 <- mean(populationn$e_0, na.rm = TRUE)
  
  summary_e0_table <- rbind(summary_e0_table, 
                            data.frame(name = pop_name, 
                                       `minimum e_0` = min_e0, 
                                       `maximum e_0` = max_e0, 
                                       `average e_0` = mean_e0))
}

kable(summary_e0_table, 
      col.names = c("Country Name", "e_0 Minimum", "e_0 Maximum","e_0 Average"),
      digits = 2,
      caption = "Female Life Expectancy at Birth in Years: Top 4 Cyclone-Prone Nations (2024) vs. World") %>%
  kable_styling(font_size = 15)

```

***(e)15 points** Choose one of the long format data.frames you just
created. Use `ggplot + geom_line()` to create a plot of female life
expectancy against time by the 5 regions (4 countries + world) in your
dataset. Print the plot in our output **AND** save this plot as a .pdf.
Comment on how the life expectancy trajectories of the four countries
you chose compare to the overall trajectory of life expectancy across
the World?*

```{r data wrangling, eval=FALSE, echo=FALSE}
# cyclone_data_long <- cyclone_data_long %>%
#   select(-country_code)
# cyclone_data_long

# cyclone_data_wide <- cyclone_data_long %>%
#   pivot_wider(names_from = name, values_from = e_0)
# 
# cyclone_data_wide
```

```{r cyclone female e0 ggplot to PDF, eval=TRUE, echo=TRUE}
# Save the plot as a PDF
pdf("cyclone_e0_plot.pdf", 
    width = 6, height = 6)

cyclone_data_long$Year <- as.numeric(cyclone_data_long$Year)

ggplot(cyclone_data_long, aes(x = Year, y = e_0, color = name)) +
  geom_line(size = 1) +
  labs(
    title = "Female Life Expectancy Over Time: Cyclone-Prone Nations vs. World",
    x = "Year",
    y = "Life Expectancy at Birth (Years)",
    color = "Country"
  ) +
  scale_color_manual(values = c("chocolate", "goldenrod", "seagreen", "steelblue", "slateblue")) +
  scale_x_continuous(breaks = seq(min(cyclone_data_long$Year), max(cyclone_data_long$Year), by = 5)) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

dev.off()
```

```{r cyclone female e0 ggplot, eval=TRUE, echo=TRUE}

cyclone_data_long$Year <- as.numeric(cyclone_data_long$Year)

ggplot(cyclone_data_long, aes(x = Year, y = e_0, color = name)) +
  geom_line(size = 1) +
  labs(
    title = "Female Life Expectancy Over Time: Cyclone-Prone Nations vs. World",
    x = "Year",
    y = "Life Expectancy at Birth (Years)",
    color = "Country"
  ) +
  scale_color_manual(values = c("chocolate", "goldenrod", "seagreen", "steelblue", "slateblue")) +
  scale_x_continuous(breaks = seq(min(cyclone_data_long$Year), max(cyclone_data_long$Year), by = 5)) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

This plot shows the scale of great dips in life expectancy at birth,
particularly in China (red) in 1960. This was due to the Great Chinese
Famine which occurred between 1959-1961 and is attributed to the Great
Leap Forward by many historians. According to
[Wikipedia](https://en.wikipedia.org/wiki/Great_Chinese_Famine#cite_note-:0-4)
(cited by many journal articles,
[this](https://pmc.ncbi.nlm.nih.gov/articles/PMC1127087/) being one of
them) this famine is considered one of the deadliest in history.

All nations except the Philippines appear to be surpassing the World's
trends, especially from about the mid 1970s onward.

# Problem 2: Histograms - 15 points

***(a)5 points** Create two histograms using base `R` of the male life
expectancies and female life expectancies in 2023 that has both
histograms side-by-side in a two-panel plot. Show your plot in your
output and save it to .pdf.*

I am assuming we can use the same four-country data as in the first
problem. These histograms will be life expectancies in 2023 of nations
experiencing the most cyclones in 2024. I am also assuming that instead
of binning the life expectancies in bins of 5-year increments, this
question is asking for a side-by-side comparison of the 2023 life
expectancies between the chosen nations + world. Therefore, instead of
making the histograms using hist() (which would require )

```{r male data wrangling, eval=TRUE, echo=FALSE}
# Male data wrangling
male_cyclone_data = e0M1 %>%
  filter(country_code %in% c(156, 608, 392, 484, 900))

male_cyclone_long <- male_cyclone_data %>%
  pivot_longer(
    cols = `1950`:`2023`,  # Specify the columns from years 1951 to 2023
    names_to = "Year",     # name of new SINGLE column for year from MANY columns
    values_to = "e_0"      # name of new column for life expectancy values
  )

# male_cyclone_long

```

```{r data prep for histograms, eval=TRUE, echo=FALSE}
# cyclone_data_long

cyclone_data_long_2023 = cyclone_data_long %>% 
  filter(Year == 2023) %>% 
  select(-Year)
male_cyclone_long_2023 = male_cyclone_long %>% 
  filter(Year == 2023) %>% 
  select(-Year)

# cyclone_data_long_2023
# male_cyclone_long_2023

```

The question asked for histograms, but I'm also adding a similar
visualization that may achieve what the question is asking better below,
using barplot().

```{r cyclone base R hist to pdf, eval=TRUE, echo=TRUE}
pdf("2023_e0_freqhistograms.pdf",
    width = 10, height = 6)

par(mfrow = c(1, 2))

hist(cyclone_data_long_2023$e_0,
  main = "Female Life Expectancies (2023)",
  xlab = "Life Expectancy in Years",
  col = "goldenrod",
  xlim = c(65, 90),
  breaks = 5
)

hist(male_cyclone_long_2023$e_0,
  main = "Male Life Expectancies (2023)",
  xlab = "Life Expectancy in Years",
  col = "seagreen",
  xlim = c(65, 90),
  breaks = 5
)

dev.off()
```

```{r cyclone base R hist, eval=TRUE, echo=TRUE}
par(mfrow = c(1, 2))

hist(cyclone_data_long_2023$e_0,
  main = "Female Life Expectancies (2023)",
  xlab = "Life Expectancy in Years",
  col = "goldenrod",
  xlim = c(65, 90),
  breaks = 5
)

hist(male_cyclone_long_2023$e_0,
  main = "Male Life Expectancies (2023)",
  xlab = "Life Expectancy in Years",
  col = "seagreen",
  xlim = c(65, 90),
  breaks = 5
)
```

```{r cyclone base R barplot to pdf, eval=TRUE, echo=TRUE}
pdf("2023_e0_countrycomparisons.pdf",
    width = 10, height = 6)

par(mfrow = c(1, 2))
barplot(
  cyclone_data_long_2023$e_0,
  names.arg = cyclone_data_long_2023$name,
  main = "Female Life Expectancies (2023)",
  xlab = "Country",
  ylab = "Life Expectancy in Years",
  col = "goldenrod",
  las = 2  
)

barplot(
  male_cyclone_long_2023$e_0,
  names.arg = male_cyclone_long_2023$name,
  main = "Male Life Expectancies (2023)",
  xlab = "Country",
  ylab = "Life Expectancy in Years",
  col = "seagreen",
  las = 2  
)

dev.off()
```

```{r cyclone base R barplot, eval=TRUE, echo=TRUE}
par(mfrow = c(1, 2))
barplot(
  cyclone_data_long_2023$e_0,
  names.arg = cyclone_data_long_2023$name,
  main = "Female Life Expectancies (2023)",
  xlab = "Country",
  ylab = "Life Expectancy in Years",
  col = "goldenrod",
  las = 2  
)

barplot(
  male_cyclone_long_2023$e_0,
  names.arg = male_cyclone_long_2023$name,
  main = "Male Life Expectancies (2023)",
  xlab = "Country",
  ylab = "Life Expectancy in Years",
  col = "seagreen",
  las = 2  
)

```

***(b)5 points** Create a histogram using `ggplot2` of the male life
expectancies and female life expectancies in 2023 that has both
histograms side-by-side in a two-panel plot. Show your plot in your
output and save it to .pdf. plot in your output and save it to .pdf.*

```{r cyclone ggplot hist side by side, eval=TRUE, echo=TRUE}

plot1 = ggplot(cyclone_data_long_2023, aes(x = name, y = e_0)) +
  geom_bar(stat = "identity", fill = "goldenrod", color = "white") +
  labs(
    title = "Female e0 by Country (2023)",
    x = "Country",
    y = "Life Expectancy in Years"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

plot2 = ggplot(male_cyclone_long_2023, aes(x = name, y = e_0)) +
  geom_bar(stat = "identity", fill = "seagreen", color = "white") +
  labs(
    title = "Male e0 by Country (2023)",
    x = "Country",
    y = "Life Expectancy in Years"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  

# Combine the plots with increased space between them
combined_plot = plot1 + plot2 + plot_layout(ncol = 2, widths = c(1, 1))  
combined_plot

ggsave("cyclone_e0_by_gender_sided.pdf", combined_plot, width = 20, height = 8)

```

***(c)5 points** Create a histogram using `ggplot2` of the male life
expectancies and female life expectancies in 2023 that has both female
and male life expectancies are presented in the same plot with different
colors. Show your plot in your output and save it to .pdf.*

```{r cyclone ggplot hist dodged, eval=TRUE, echo=TRUE}

female_data = cyclone_data_long_2023 %>% 
  mutate(Gender = "Female")
male_data <- male_cyclone_long_2023 %>% 
  mutate(Gender = "Male")

combined_data = bind_rows(female_data, male_data)

ggplot(combined_data, aes(x = name, y = e_0, fill = Gender)) +
  geom_bar(stat = "identity", 
           position = "dodge", 
           color = "white") +
  labs(
    title = "Life Expectancy by Country and Gender (2023)",
    x = "Country",
    y = "Life Expectancy in Years"
  ) +
  scale_fill_manual(values = c("goldenrod", "seagreen")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave('cyclone_e0_by_gender_dodge.pdf')
```

# Problem 3: Barplots - 15 points

*Load the dataset `UNlocations` from the `wpp2024` package and read the
documentation.*

Description: United Nations table of locations, including regions, for
statistical purposes as available in 2024.

Source: Data provided by the United Nations Population Division.

Note: the agcode columns are optional columns that can be used for
aggregations. To aggregate a region with country_code=x, get the value
of its location_type, say y. Then, look for the column agcode_y and
locate all records with agcode_y=x that have location_type=4.

```{r unloc data load, eval=TRUE, echo=FALSE}

data(`UNlocations`)

# print(colnames(UNlocations))
# summary(UNlocations)
# head(UNlocations)

# Remove AgCode columns (optional and not necessary for this lab)
UNlocations_raw = UNlocations %>%
  select(-contains("agcode"))
# UNlocations_raw

```

***(a)5 points** Create a barplot in base `R` that has a bar for each
region of the world where the height of the bars is the number of
countries in that region.*

I am assuming the granularity for regions of the world that would best
reflect my barplot's data would be reg_name because it is more granular
than area_name. For example, area_name refers to countries in 'Western
Asia' and 'South-Eastern Asia' as just 'Asia'.

```{r region counting, eval=TRUE, echo=FALSE}
# world_regions = UNlocations_raw$name[2:9]
# world_regions

countries_by_area = table(UNlocations_raw$reg_name)

# Remove the first column because it's a set of all the rows which are categories, such as "Africa" or "High Income Countries", not countries.
countries_by_area = countries_by_area[-1]

# max(countries_by_area)
```

```{r base R barplot of regions by count, eval=TRUE, echo=TRUE}

barplot(
  countries_by_area,
  main = "Countries by Region",
  xlab = "Regions of the World",
  ylab = "Number of Countries",
  col = "orchid4",
  # so that the labels all fit
  las = 2,           
  # so that they're legible
  cex.names = 0.4, 
  # unless the ytick labels will all auto-set to max 25
  ylim = c(0, 30)    
)

```

***(b)5 points** Create a barplot in `ggplot2` that has a bar for each
region of the world where the height of the bars is the number of
countries in that region.*

```{r ggplot of regions by count, eval=TRUE, echo=TRUE}
# base R bar plots don't work so well with dataframes but ggplot does
countries_by_area_df <- as.data.frame(countries_by_area)

ggplot(countries_by_area_df, aes(x = Var1, y = Freq)) + 
  geom_bar(stat = "identity", fill = "orchid4", color = "white") +
  labs(
    title = "Countries by Region",
    x = "Regions of the World",
    y = "Number of Countries"
  ) +
  theme_minimal() + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 6),
    axis.text.y = element_text(size = 12),
    plot.title = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    strip.text.x = element_text(size = 10)
  ) +
  scale_y_continuous(breaks = seq(0, 30, by = 5))
```

***(c)5 points** For at least one of the above plots color each bar a
different color.*

```{r viridis barplot of regions by count, eval=TRUE, echo=TRUE}
ggplot(countries_by_area_df, aes(x = Var1, y = Freq, fill = Var1)) + 
  geom_bar(stat = "identity", color = "white") +
  labs(
    title = "Countries by Region",
    x = "Regions of the World",
    y = "Number of Countries"
  ) +
  theme_minimal() + 
  theme(legend.position = "none")  +  
  #use this last line to get rid of the legend 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 6),
    axis.text.y = element_text(size = 12),
    plot.title = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    strip.text.x = element_text(size = 10)
  ) +
  scale_y_continuous(breaks = seq(0, 30, by = 5)) +
  # this is the most common one I've been taught and I think it's also color blind accessible
  scale_fill_viridis_d() 

```

# Problem 4: Merging Two Datasets - 20 points

*Check out the documentation for the base `R` function `merge` and the
dplyr functions `left_join` and `full_join`.*

```{r merge, eval=TRUE, echo=FALSE}
?merge
```

Description: Merge two data frames by common columns or row names, or do
other versions of database join operations.

Usage: merge(x, y, ...)

merge(x, y, by = intersect(names(x), names(y)), by.x = by, by.y = by,
all = FALSE, all.x = all, all.y = all, sort = TRUE, suffixes =
c(".x",".y"), no.dups = TRUE, incomparables = NULL, ...)

```{r left join, eval=TRUE, echo=FALSE}
?left_join
```

Description: Mutating joins add columns from y to x, matching
observations based on the keys. There are four mutating joins: the inner
join, and the three outer joins.

Inner join: An inner_join() only keeps observations from x that have a
matching key in y. This means that generally inner joins are not
appropriate in most analyses, because it is too easy to lose
observations.

Outer joins: The three outer joins keep observations that appear in at
least one of the data frames:

A left_join() keeps all observations in x.

A right_join() keeps all observations in y.

A full_join() keeps all observations in x and y.

left_join( x, y, by = NULL, copy = FALSE, suffix = c(".x", ".y"), ...,
keep = NULL )

left_join( x, y, by = NULL, copy = FALSE, suffix = c(".x", ".y"), ...,
keep = NULL, na_matches = c("na", "never"), multiple = "all", unmatched
= "drop", relationship = NULL )

```{r full join, eval=TRUE, echo=FALSE}
?full_join
```

full_join( x, y, by = NULL, copy = FALSE, suffix = c(".x", ".y"), ...,
keep = NULL )

full_join( x, y, by = NULL, copy = FALSE, suffix = c(".x", ".y"), ...,
keep = NULL, na_matches = c("na", "never"), multiple = "all",
relationship = NULL )

***(a)10 points** Use one of these functions to create a new data.frame
called `e0` that contains all information from `e0F1` and `e0M1`. Pivot
the data.frame longer to create an object called `e0_long` so that there
is a new column called `Sex` indicating whether the row contains life
expectancies for "Female" or "Male".*

Because the question is asking for all information from `e0F1` and
`e0M1` I will use full_join before pivoting.

I am including information about using regex to grab the year columns
affected by the pivot by using the following, but not using it in the
code below:

1.  \^: Matches the start of the string.
2.  [0-9]: Matches any digit (0–9).
3.  {4}: Matches exactly 4 digits (e.g., 1950, 2023).
4.  \\.: Matches a literal period (.). The \\ is needed to escape the
    '.' in regex, as '.' is like a wild card and can match any
    character.
5.  (x\|y): Matches either x or y. This is a group using the \| operator
    (OR).
6.  \$: Matches the end of the string.

It's easier to use the select function because it is convenient, but
that is also more hard-coded than using regex, so I'm including it as a
comment in the code in case it's useful to me later.

```{r e0 join, eval=TRUE, echo=TRUE}
# Full Join the two dataframes together
e0 <- full_join(e0F1, e0M1, by = "country_code")
e0 = e0 %>% select(-name.y)
e0 = e0 %>% rename(name = name.x) 
# e0
```

The reason I chose not to present the entire e0_long dataframe, even
with a Kable-style table, is because it is so large that the document
fails to render.

```{r e0_long pivot, eval=TRUE, echo=TRUE}
e0_long <- e0 %>%
  pivot_longer(
    # cols = matches("^[0-9]{4}\\.(x|y)$") would also work
    cols = -c(country_code, name), 
    names_to = c("year", "sex"),
    names_sep = "\\.",
    values_to = "e_0"
  )%>%
  # Map 'x' and 'y' to 'Female' and 'Male'
  mutate(
    sex = ifelse(sex == "x", "Female", "Male") 
  )

# nrow(e0_long) #sanity check
for_show = head(e0_long, 51)

#present it
kable(for_show,
      col.names = c("Country Code", "Name", "Year", "Sex" , "Life Expectancy at Birth"),
      digits = 2,
      caption = "Long Format of e0 data in 20th Century: Male & Female Life Expectancy at Birth") %>%
  kable_styling(
    font_size = 15
  ) %>%
  scroll_box(width = "900px", height = "500px")
```

***(b)5 points** Pick one country from those available and make a new
data.frame called `e0.small` that contains only rows with data from
those country.*

I am choosing Cambodia with country code = 116. I read a paper in the
previous demography-track course on new methodology for estimating
excess deaths during the PPR (Pol Pot Regime). Those researchers used
survivor interviews, satellite imagery from mass graves, and voter
records, among other avenues, to make an educated guess. I don't have
any experience seeing how this affects aggregate life expectancy data so
I thought this would be a good opportunity to learn.

```{r e0_small, eval=TRUE, echo=TRUE}

e0_small = e0_long %>% 
  filter(country_code == 116)

# dim(e0_small)
# e0_small

kable(e0_small,
      col.names = c("Country Code", "Name", "Year", "Sex" , "Life Expectancy at Birth"),
      digits = 2,
      caption = "Long Format of Male & Female Life Expectancy at Birth in Cambodia") %>%
  kable_styling(
    font_size = 15
  ) %>%
  scroll_box(width = "900px", height = "500px")

```

***(c)5 points** Plot the male and female trajectories of life
expectancy for that country using base `R` and `ggplot2`, using either
color or line type to distinguish between the trajectories.*

Base R plotting code:

```{r base r Cambodia, eval=TRUE, echo=TRUE}

cambodia_longF = e0_small[e0_small$sex == "Female", ]
cambodia_longM <- e0_small[e0_small$sex == "Male", ]

plot(cambodia_longF$year, cambodia_longF$e_0, 
     type = "l", 
     col = "goldenrod", 
     lwd = 3, 
     xlab = "Year", 
     ylab = "Life Expectancy (e_0)", 
     main = "Life Expectancy Trajectories (Cambodia)",
     ylim = range(e0_small$e_0),
     xaxt = "n")

years = seq(1950, 2025, by = 5)
axis(1, at = years, labels = years)

lines(cambodia_longM$year, cambodia_longM$e_0, 
      col = "seagreen", 
      lwd = 3)

#grid lines
grid(nx = NULL, ny = NULL, col = "lightgray")

legend("bottomright", 
       legend = c("Female", "Male"), 
       col = c("goldenrod", "seagreen"), 
       lwd = 3)

```

ggplot2 plotting code:

```{r ggplot Cambodia, eval=TRUE, echo=TRUE}
e0_small$year = as.numeric(e0_small$year)

ggplot(e0_small, aes(x = year, y = e_0, color = sex)) +
  annotate("rect", xmin = 1975, xmax = 1979, ymin = -Inf, ymax = Inf, 
           fill = "red", alpha = 0.25) +
  geom_line(size = 1) +
  labs(
    title = "Life Expectancy Trajectories (Cambodia)",
    x = "Year",
    y = "Life Expectancy (e_0)",
    color = "Sex"
  ) +
  scale_color_manual(
    values = c("Female" = "goldenrod", "Male" = "seagreen")
  ) +
  scale_x_continuous(breaks = seq(min(e0_small$year), max(e0_small$year), by = 5)) +
  scale_y_continuous(breaks = seq(0, 80, by = 5)) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

The PPR period lasted from 1975-1979.
